# Visualización de mapas basados en la web

## Objetivos

En esta clase:

- Como usar RMarkdown para crear un sitio web
- Una revisión rápida de como producir visualizaciones con R y el paquete `ggplot`
- Como creat mapas interactivos usando el paquete `leaflet`
- Piblicar mapas interactivos usando RMarkdown para hacer un sitio web en GitHub

## Introducción

Compartir el trabajo científico de una manera interativa y entertenida es fundamental para el 
desarrollo del proceso científico. Hasta este punto hemos hablado en este curso sobre un set
}de herramientas que nos ayudan a hacer ciencia abierta (_open science_):

- R y muchos de sus paquetes
- RStudio
- git
- GitHub
- RMarkdown

RMarkdown, en particular, es increiblemente potente para crear reportes científicos, sin embargo
hasta ahora no hemos llegado ni a rozar en su verdaro potencial para compartir nuetro trabajo
con colaboradores.

En esta leccion, vamos a tomar un repositorio ya existente en GitHub y lo transformaremos en un 
sitio web elegante y simple de leer usando algunas de las herramientas que hemos aprendido en este 
curso.

## Un ejemplo básico

- En RStudio, cree im archivo nuevo al nivel tope se su repositorio **git** llamado `index.Rmd`. 
La forma más simple de hacer esto es a través del menú de RStudio. Elija File -> New File -> RMarkdown...
Esto abrirá una caja de dialogo. Usted debe crear un "Document" en formato "HTML". Estas son las 
opciones por defecto.
- Abra `index.Rmd` (si es que no está y abierto)
- Haga clic en _Knit_
    - Vea la salida que se genera
    - Note que se ha creado un nuevo archivo `index.html` en el mismo directorio. Este es el resultado
    en HTML (página web) creado a partir del documento RMarkdown 
- Commit sus cambios (ambos archivos, index.Rmd e index.html)
- Abra su navegador web en la página www.GitHub.com de su repositorio
- Vaya a Settings > GitHub Pages y active _GitHub Pages_ para la rama (branch) `master`

    Ahora la versión del sitio web rendered de su repositorio se va a mostrar en un URL special.
    
    GitHub Pages sigue una convensión como esta:
    
     ```
     https://{username}.github.io/{repository}/
     https://cornejotux.github.io/claseRMarkdown/
     ```
     
    Ponga atención que ahora sitio no está en github.com si no que en github.io. 
     
- Vaya a https://{username}.github.io/{repo_name}/ (Hay que poner el `/` al final!)
    Y vea la versión generada de su sitio web.
    

## Un ejemplo ya no tan básico

Now that we've seen how to create a web page from RMarkdown, let's create a website that uses some of the cool functionality available to us.
We'll use the same git repository and RStudio Project as above, but we'll be adding some files to the repository and modifying `index.Rmd`.

First, let's get some data. We'll re-use the salmon escapement data from the ADF&G OceanAK database:

- Navigate to [Escapement Counts](https://doi.org/10.5063/F1S46Q6M) (or visit the KNB and search for 'oceanak') and copy the Download URL for the `ADFG_firstAttempt_reformatted.csv` file
- Download that file into R using `read.csv` to make the script portable
- Calculate annual escapement by species and region using the `dplyr` package
- Make a bar plot of the annual escapement by species using the `ggplot2` package
- Display it in an interactive table with the `datatable` function from the `DT` package
- And lastly, let's make an interactive, Google Maps-like map of the escapement sampling locations.

To do this, we'll use the [`leaflet`](https://rstudio.github.io/leaflet/) package to create an interactive map with markers for all the sampling locations:

First, let's load the packages we'll need:

```{r load_packages, message = F, warning = F}
  library(leaflet)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(DT)
  library(sf)
  library(ggmap) # devtools::install_github("dkahle/ggmap")
```

### Carga de datos sobre retorno de salmones

You can load the data table directly from the KNB Data Repository, if it isn't
already present on your local computer.  This technique 

```{r read_data}
data_url <- "https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Af119a05b-bbe7-4aea-93c6-85434dcb1c5e"

esc <- tryCatch(
    read.csv("data/escapement.csv", stringsAsFactors = FALSE),
    error=function(cond) {
        message(paste("Escapement file does not seem to exist, so get it from the KNB."))
        esc <- read.csv(url(data_url, method = "libcurl"), stringsAsFactors = FALSE)
        return(esc)
    }
)

head(esc)
```

### Static Plots

Now that we have the data loaded, let's calculate annual escapement by species and region:

```{r mean_esc}
annual_esc <- esc %>% 
  separate(sampleDate, c("Year", "Month", "Day"), sep = "-") %>% 
  mutate(Year = as.numeric(Year)) %>% 
  group_by(Species, SASAP.Region, Year) %>% 
  summarize(escapement = sum(DailyCount)) %>% 
  filter(Species %in% c("Chinook", "Sockeye", "Chum", "Coho", "Pink"))

head(annual_esc)
```

That command used a lot of the dplyr commands that we've used, and some that are new.
The `separate` function is used to divide the sampleDate column up into Year, Month, and Day columns,
and then we use `group_by` to indicate that we want to calculate our results for
the unique combinations of species, region, and year.  We next use `summarize`
to calculate an escapement value for each of these groups. Finally, we use a filter and the `%in%` operator to select only the salmon species.

Now, let's plot our results using `ggplot`. `ggplot` uses a mapping *aesthetic* (set using `aes()`) and a *geometry* to create your plot. Additional geometries/aesthetics and theme elements can be added to a `ggplot` object using `+`.

```{r plot_esc}
ggplot(annual_esc, aes(x = Species, y = escapement)) +
  geom_col()
```

What if we want our bars to be blue instad of gray? You might think we could run this:

```{r}
ggplot(annual_esc, aes(x = Species, y = escapement, fill = "blue")) +
  geom_col()
```

Why did that happen?

Notice that we tried to set the fill color of the plot inside the mapping aesthetic call. What we have done, behind the scenes, is create a column filled with the word "blue" in our dataframe, and then mapped it to the `fill` aesthetic, which then chose the default fill color of red. 

What we really wanted to do was just change the color of the bars. If we want do do that, we can call the color option in the `geom_bar` function, **outside** of the mapping aesthetics function call.

```{r}
ggplot(annual_esc, aes(x = Species, y = escapement)) +
  geom_col(fill = "blue")
```

What if we did want to map the color of the bars to a variable, such as region.

`ggplot` is really powerful because we can easily get this plot to visualize more aspects of our data.

```{r}
ggplot(annual_esc, aes(x = Species, y = escapement, fill = SASAP.Region)) +
  geom_col()
```

Here is an example of a different aesthetic and geometry mapping we can create with this data. You can also add a title, adjust labels, and include a built in theme.

```{r}
ggplot(filter(annual_esc, SASAP.Region == "Kodiak"), aes(x = Year, y = escapement, color = Species)) + 
    geom_line() +
    geom_point() +
    ylab("Escapement") +
    ggtitle("Kodiak Salmon Escapement") +
    theme_bw()

```

Note how I added a filter call in my `ggplot` function call to only plot data from Kodiak. You can easily create plots like this for every region using `facet_wrap`

```{r}
ggplot(annual_esc, aes(x = Year, y = escapement, color = Species)) + 
    geom_line() +
    geom_point() +
    facet_wrap(~SASAP.Region, scales = "free_y") +
    ylab("Escapement") +
    theme_bw()

```

### Interactive Maps

Now let's convert the escapement data into a table of just the unique locations:

```{r uniq_locations}
locations <- esc %>% 
  distinct(Location, Latitude, Longitude) %>% 
  drop_na()
```

And display it as an interactive table:

```{r display_locations}
datatable(locations)
```

Then making a leaflet map is (generally) only a couple of lines of code:

```{r simple_map}
leaflet(locations) %>% 
  addTiles() %>% 
  addMarkers(~ Longitude, ~ Latitude, popup = ~ Location)
```

The `addTiles()` function gets a base layer of tiles from [OpenStreetMap](https://www.openstreetmap.org/) which is an open alternative to Google Maps.
`addMarkers` use a bit of an odd syntax in that it looks kind of like `ggplot2` code but uses `~` before the column names.
This is similar to how the `lm` function (and others) work but you'll have to make sure you type the `~` for your map to work.

You can also use `leaflet` to import Web Map Service (WMS) tiles. Here is an example that utilizes the General Bathymetric Map of the Oceans [(GEBCO)](https://www.gebco.net/data_and_products/gebco_web_services/web_map_service/#getcapabilities) WMS tiles. In this example, we also demonstrate how to create a more simple circle marker.

```{r}
leaflet(locations) %>% 
  addWMSTiles("https://www.gebco.net/data_and_products/gebco_web_services/web_map_service/mapserv?",
              layers = 'GEBCO_LATEST',
              attribution = "Imagery reproduced from the GEBCO_2014 Grid, version 20150318, www.gebco.net") %>%
  addCircleMarkers(lng = ~Longitude,
                   lat = ~Latitude,
                   popup = ~ Location,
                   radius = 5,
                   # set fill properties
                   fillColor = "salmon",
                   fillOpacity = 1,
                   # set stroke properties
                   stroke = T,
                   weight = 0.5,
                   color = "white",
                   opacity = 1)
```

Leaflet has a ton of functionality that can enable you to create some beautiful, functional maps with relative ease. [Here](https://pages.github.nceas.ucsb.edu/NCEAS/sasap-data/language_vis.html) is an example of some we created as part of the SASAP project, created using the same tools we showed you here. This map hopefully gives you an idea of how powerful the combination of RMarkdown and GitHub pages can be. 


## Static maps with geom_sf

We can also create a similar static map using `ggplot` with the `geom_sf` geometry. To
do so, first we need to transform our locations data frame into an `sf` object by
providing the coordinate reference system (EPSG:4326 is the CRS for geocoordinates in WGS84).
We also transform the points to `EPSG:3857` which is the CRS used for rendering 
maps in Google Maps, Stamen, and OpenStreetMap, among others.

```{r}
locations_sf <- locations %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
locations_sf_3857 <- st_transform(locations_sf, 3857)
```

Next, let's grab a base map from the Stamen map tile server covering the region of interest.
This includes a fix to transform the bounding box (which starts in `EPSSG:326`) to also be in the `EPSG:3857` CRS,
which is the projection that the map raster is returned in from Stamen.


```{r message=FALSE}
# Define a function to fix the bbox to be in EPSG:3857
# See https://github.com/dkahle/ggmap/issues/160#issuecomment-397055208
ggmap_bbox_to_3857 <- function(map) {
  if (!inherits(map, "ggmap")) stop("map must be a ggmap object")
  # Extract the bounding box (in lat/lon) from the ggmap to a numeric vector, 
  # and set the names to what sf::st_bbox expects:
  map_bbox <- setNames(unlist(attr(map, "bb")), 
                       c("ymin", "xmin", "ymax", "xmax"))
  
  # Coonvert the bbox to an sf polygon, transform it to 3857, 
  # and convert back to a bbox (convoluted, but it works)
  bbox_3857 <- st_bbox(st_transform(st_as_sfc(st_bbox(map_bbox, crs = 4326)), 3857))
  
  # Overwrite the bbox of the ggmap object with the transformed coordinates 
  attr(map, "bb")$ll.lat <- bbox_3857["ymin"]
  attr(map, "bb")$ll.lon <- bbox_3857["xmin"]
  attr(map, "bb")$ur.lat <- bbox_3857["ymax"]
  attr(map, "bb")$ur.lon <- bbox_3857["xmax"]
  map
}

bbox <- c(-170, 52, -130, 64)   # This is roughly southern Alaska
ak_map <- get_stamenmap(bbox, zoom = 4)
ak_map_3857 <- ggmap_bbox_to_3857(ak_map)
```

 Finally, plot both the base raster map with the points overlayed, which is easy now
 that everything is in the same projection (3857):

```{r message=FALSE}
ggmap(ak_map_3857) + 
  geom_sf(data = locations_sf_3857, inherit.aes = FALSE)
```

## Resources

- Lisa Charlotte Rost. (2018) [Why not to use two axes, and what to use instead: The case against dual axis charts](https://blog.datawrapper.de/dualaxis/)
